PyTracer: Automatically Profiling Numerical Instabilities in Python

### 2022-05-16 IEEEtran/paper.tex
More edits

\remove{Listings~\mbox{\ref{fig:wrapper_creation}} and~\mbox{\ref{fig:generic_wrapper}} show
    \mbox{\pytracer}'s function instrumentation mechanism.}
For each function to instrument, \pytracer's \texttt{loader} class creates a
wrapper function, dynamically compiles it with the \texttt{compile} builtin
function, and substitutes it for the original module function.  Although simple,
this technique does not apply to callable class instances, i.e., class instances
\add{that can be called directly with the () operator, because they,} are not of type
\texttt{FunctionType}. Indeed, applying the wrapper technique would 1) modify
the type of the class instance to \texttt{FunctionType}, which could cause
syntax and semantic errors, and 2) mask class attributes, leading to
\texttt{AttributeError} exceptions. \pytracer overrides the
\texttt{\_\_call\_\_} attribute with the wrapper function when possible to
overcome this issue. When the \texttt{\_\_call\_\_} attribute is not writable,
Pytracer does not instrument the class and returns a warning.

@@

In functional programming, iterators traverse containers lazily, meaning that
the next element in the sequence is only computed when the application uses it.
This technique allows for the manipulation of virtually infinite sequences with
finite memory. However, it implies that no complete mapping of the container
returned by an iterator exists in memory since the application computes each
element on the fly. Therefore, iterators are not serializable which implies that
\pytracer cannot save them to output traces. A workaround would be to convert
iterators into explicit containers, however, this would increase the memory
footprint significantly. \pytracer, therefore, does not instrument non-builtin
iterators.

@@

Once traces are available, \pytracer sequentially parses them and computes the
mean, standard deviation, and the number of significant digits (as in
Equation~\ref{eq:sig-digits}) of all saved values. \pytracer saves these summary
statistics in an HDF5 file, a hierarchical format that facilitates visual
browsing since only required information are loaded when requested by the user.
This operation critically relies on the ordering of call input and output
objects in the traces. Therefore, it assumes that the application is
single-threaded and that the control flow does not depend on a random state.
\add{Therefore, \mbox{\pytracer} cannot merge traces resulting from different
    control flow paths and will stop the parsing when encountering such a situation.}

@@

The second example compares a Bayesian Ridge Regression (BRR) to the Ordinary
Least Squares (OLS) estimator on a synthetic dataset and for one-dimensional
regression using polynomial feature expansion. Although the example does not
raise a runtime error, the regression coefficients computed from the fitting are
non-significant. \pytracer traces reveal that the SVD solver used is once again the root
cause for this error. \add{Indeed, the BayesianRidge model first computes an SVD
    before iteratively updating the coefficients until reaching the convergence
    criterion. However, PyTracer shows an issue in the singular values computed.}
\add{We can see in Figure}~\ref{fig:BRR_svd} \add{that the last singular values is
    \texttt{NaN} due to numerical errors in the SVD, although the remaining
    singular values exhibit high \add{result} precision (between 44 and 50 bits).
    Surprinsingly, this \texttt{NaN} value does not raise an exception and is silently
    propagated. Indeed, the stopping criterion is never verified and the 
    optimization stops after reaching the maximal number of iterations.
    Figure}~\ref{fig:BRR_svd_fixed} \add{shows that uses
    the \texttt{gesvd} instead of the \texttt{gesdd} method removes the \texttt{NaN} value
    although the last singular values stills have a low \add{result} precision (below 0).
    Following Equation}~\ref{eq:sig-digits}, \add{a number of significant digits below 0 means that the
    variability ($\sigma$) is higher than the average value ($\mu$)}.

@@

\subsection{\pytracer instrumentation limitations}

\add{
    The limitations of the PyTracer's instrumentation have little impact in practice
    since we have encountered few classes having non-writeable attributes in practice.
    Moreover, those classes are generally used in higher-level classes written in
    pure Python. For instance, the SGD class that implements the Stochastic Gradient
    Descent in scikit-learn takes a loss function as a parameter. Loss functions are
    implemented in Cython in scikit-learn and Pytracer is not able to instrument
    them. However, Pytracer can dump the instance of LossFunction if they are used
    by the SGD class as arguments, which is a frequently observed pattern in
    Scikit-learn for instance. So one still has an estimation of variability for the
    attributes of the class. Finally, one can use Veritracer to trace the methods of
    those Cython classes for an in-depth analysis since it is the C language
    underhood. We have encountered few iterators and none of them involved floating
    point values so we were able to safely ignore them. 
}

@@

In the future, we plan to extend the post-processing analysis to aggregate
traces \add{obtained with different control flow paths,} by leveraging techniques used in
code coverage analysis. Moreover, we would like to extend \pytracer to
instrument elementary arithmetic operations. \remove{Finally, although \pytracer supports multiple noise models,
we only tested it with Monte-Carlo arithmetic. Therefore, we would like to
compare different models against MCA.} \pytracer is available at
\href{https://github.com/big-data-lab-team/pytracer}{github://big-data-lab-team/pytracer}
under MIT License.

>>>>

\remove{Listings~\mbox{\ref{fig:wrapper_creation}} and~\mbox{\ref{fig:generic_wrapper}} show
    \mbox{\pytracer}'s function instrumentation mechanism.}
For each function to instrument, \pytracer's \texttt{loader} class creates a
wrapper function, dynamically compiles it with the \texttt{compile} builtin
function, and substitutes it for the original module function.  Although simple,
this technique does not apply to callable class instances, i.e., class instances
\add{that can be called directly with the () operator, because they,} are not of type
\texttt{FunctionType}. Indeed, applying the wrapper technique would 1) modify
the type of the class instance to \texttt{FunctionType}, which could cause
syntax and semantic errors, and 2) mask class attributes, leading to
\texttt{AttributeError} exceptions. \pytracer overrides the
\texttt{\_\_call\_\_} attribute with the wrapper function when possible to
overcome this issue. When the \texttt{\_\_call\_\_} attribute is not writable,
Pytracer does not instrument the class and returns a warning. 
\add{
    This limitation is expected to have a minor impact since non-writable
    classes are usually called from pure Python classes that can be
    instrumented. PyTracer reports information about the stability of
    non-writable classes through the instrumentation of the pure Python
    classes that use them.
}

@@

In functional programming, iterators traverse containers lazily, meaning that
the next element in the sequence is only computed when the application uses it.
This technique allows for the manipulation of virtually infinite sequences with
finite memory. However, it implies that no complete mapping of the container
returned by an iterator exists in memory since the application computes each
element on the fly. Therefore, iterators are not serializable which implies that
\pytracer cannot save them to output traces. A workaround would be to convert
iterators into explicit containers, however, this would increase the memory
footprint significantly. \pytracer, therefore, does not instrument
iterators.

\add{
We do not expect this limitation to have a critical impact on the
applicability of \pytracer. 
In general, iterators involve simple computations that are unlikely to be
unstable. In particular, we encountered only a few iterators in our
experiments and none of them involved floating point values. Should
instabilities occur in iterators, \pytracer would detect them through the
inputs of the functions consuming these iterators. 
}

@@

Once traces are available, \pytracer sequentially parses them and computes the
mean, standard deviation, and the number of significant digits (as in
Equation~\ref{eq:sig-digits}) of all saved values. \pytracer saves these summary
statistics in an HDF5 file, a hierarchical format that facilitates visual
browsing since only required information are loaded when requested by the user.
This operation critically relies on the ordering of call input and output
objects in the traces. It assumes that the application is
single-threaded and that the control flow does not depend on a random state.
\add{Therefore, \mbox{\pytracer} cannot merge traces resulting from different
    control flow paths and will stop the parsing when encountering such a situation.}

@@

The second example compares a Bayesian Ridge Regression (BRR) to the Ordinary
Least Squares (OLS) estimator on a synthetic dataset and for one-dimensional
regression using polynomial feature expansion. Although the example does not
raise a runtime error, the regression coefficients computed from the fitting are
non-significant. \pytracer traces reveal that the SVD solver used is once again the root
cause for this error. \add{Indeed, the BayesianRidge model first computes an SVD
    before iteratively updating the coefficients until reaching the convergence
    criterion. However, PyTracer shows an issue in the singular values computed.}
\add{We can see in Figure}~\ref{fig:BRR_svd} \add{that the last singular values is
    \texttt{NaN} due to numerical errors in the SVD, although the remaining
    singular values exhibit high \add{result} precision (between 44 and 50 bits).
    Surprinsingly, this \texttt{NaN} value does not raise an exception and is silently
    propagated. Indeed, the stopping criterion is never verified and the 
    optimization stops after reaching the maximal number of iterations.
    Figure}~\ref{fig:BRR_svd_fixed} \add{shows that uses
    the \texttt{gesvd} instead of the \texttt{gesdd} method removes the \texttt{NaN} value
    although the last singular values stills have a low \add{result} precision (below 0).
    According to Equation}~\ref{eq:sig-digits}, \add{a number of significant bits less than 0 means that the
    variability ($\sigma$) is higher than the average value ($\mu$)}.

@@

@@

In the future, we plan to extend the post-processing analysis to aggregate
traces \add{obtained with different control flow paths,} by leveraging
techniques used in code coverage analysis. \add{This would also enable a larger support 
of multi-threaded applications.} Moreover, we would like to
extend \pytracer to instrument elementary arithmetic operations \add{(+,-,*,/), to allow for more detailed analyses}.
\remove{Finally, although \pytracer supports multiple noise models, we only
tested it with Monte-Carlo arithmetic. Therefore, we would like to compare
different models against MCA.} \pytracer is available at
\href{https://github.com/big-data-lab-team/pytracer}{github://big-data-lab-team/pytracer}
under MIT License.
